---
title: "R in an Hour"
author: "Bill Kable"
date: "June 28, 2015"
output: ioslides_presentation
---

## What are we going to cover

- Highlights of R language, a bit of history and its characteristics
- Really good areas to consider R language and eco-system
- Demo/Walkthrough of Real R projects
- We WILL NOT cover deep language syntax (sorry engineers, separate Talk ;)

## R - What is it? Where did it come from?

- Historically a DSL for Statistic analysis and reporting
- Based from S language

    - S in 1970s
    - R dialect of S in late 1990's
        
- Implemented in C and FORTAN
- Heavy adoption in Data Science in recent years
- Heavy community support to extend R core with packages - CRAN
- Competition from Python

## R - Language Characteristics
- Dynamic Typing
- Interpreted
- Functional, with some OO littered in
- Declarative vs Imperative -- emphasis on set operations
- No scalars, typing built from Vectors, Matrix, Lists and Data Frames/Data Tables
- Potentially large learning curve for Java Dev vs. Python

## R - Goodies
- Vectors

```{r vector.example, message=FALSE,warning=FALSE}
someVector <- c(1,2,3)
```

- Matrix

```{r matrix.example, message=FALSE,warning=FALSE}
someMatrix <- c(1,2,3,4,5,6, nrow=2,ncol=3)
```

- List

```{r list.example, message=FALSE,warning=FALSE}
someList <- list(someMatrix,someVector)
```

## R - Data Frame (Table)
- In R we can create tables (similar to tables in RDBMS)
- Create a Data Frame (Table)

```{r data.table.example, message=FALSE,warning=FALSE}
library(data.table)
dt <- data.table(id = c(1,2,2,3,3,3), value = c(100,200,200,300,300,300))
```

## R - Accessing element from data table
- To access an element, we can use "subsetting"
- By Row:
```{r subset.row.example, message=FALSE,warning=FALSE}
dt[2]
```
- Columns by name:
```{r subset.name.example, message=FALSE,warning=FALSE}
dt$value[2]
```

## R - Filtering data tables
- Let's filter table rows with ID of two via "subsetting"
- Subsetting is R core feature to select

```{r subset.example, message=FALSE,warning=FALSE}
dt[dt$id == 2]
```
- Or, we can use the dplyr (or plyr) filter method...
```{r filter.example, message=FALSE,warning=FALSE}
library(dplyr)
filter(dt,id == 2)
```

## R - Chaining
- Chain several aggregrate operations
```{r chain.example, message=FALSE,warning=FALSE}
filter(dt, id > 1) %>%
        group_by(id) %>% summarize(count = n(),
                                   total = sum(value))
```

## R - Key Language Takeaway: Avoid Looping Constructs!
- R Language Performance - traditional procedural looping constructs not as efficient as set based constructs
- Much effort by community to build performant set based packages
